Mergeable Types

### 2017-02-28 paper/intro.tex

To illustrate how these kinds of problems manifest in practice,
consider how we might write a simple counter library (see
Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
operations - \C{add} and \C{mult} - that lets a non-negative integer
value be added or multiplied to the counter, resp.  Observe that the
library is written in an idiomatic functional style, with no special
reasoning principles needed to realize desired functionality.  As long
as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior can become significantly harder to understand.  In
particular, a distributed implementation might wish to
\emph{replicate} the counter state on each node to improve response
time or fault tolerance.  Unfortunately, adding replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose significant performance penalties.  But, without such
heavyweight mechanisms, applying an \C{Add} operation on one node may
not be instanteously witnessed on another, which may be in the process of
simultaneously attempting to perform its own \C{Add} or \C{Mult}
action.  While synchronizing the activities of all nodes to ensure at
most one such operation is performed at a time is impractical,
designating a single node to hold the counter state eschewing
replication altogether (as in a client-server
configuration~\cite{Armstrong}), is also an undesirable solution,
given the sensitivity of such architectures to network partitions and server failures, and
the negative performance impact it incurs in geo-distributed
environments~\cite{Walter}.  Removig coordination altogether by simply
replicating counter state without having any supporting consistency
protocol is an equally infeasible approach.

@@

nonetheless witness anamolous states that violate the invariant

@@

Framing replication as merging leads to a counter implementation that
bears strong similarity to the original sequential one:
  \begin{ocaml}
    module Replicated_Counter = struct
      include Counter
      let merge lca v1 v2 =
         lca + (v1 - lca) + (v2 - lca)
    end
  \end{ocaml}
The role of \C{lca} (least common ancestor) here captures salient
history - the state resulting from the merge of two versions derived
from the same ancestor state should not unwittingly duplicate the
contributions of the ancestor.  This interpretation of a replicated
datatype is thus given in terms of the evolution of a program state
impicitly associated with the different nodes that comprise a
distributed application with merge operations serving to communicate
and reconcile different local states.

@@

The remainder of the paper is organized as follows.  We present the
\name\ programming model informally in the next section.
Sec.~\ref{sec:mergeable_types} develops a series of examples that
illustrate how merge operations can be defined and used.
Sec.~\ref{sec:formalization} defines an operational semantics, and
formalizes our intution of a mergeable type in terms of morphisms over
datatype operations.  We also present soundness (consistency) and
progress guarantees enjoyed by the
semantics. Sec.~\ref{sec:system-model} considers an instantiation of
the model suitable for distributed environments with network
partitions and failures.  Implementation details describing how we
incorporate mergeable types into OCaml are presented in
Sec.~\ref{sec:impl}.  Experimental results and benchmarks are given in
Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related}
and~\ref{sec:conclusions} describes related work and conclusions.

>>>>

To illustrate how these kinds of problems manifest in practice,
consider how we might write a simple counter library (see
Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
operations - \C{add} and \C{mult} - that lets a non-negative integer
value be added or multiplied to the counter, resp.  Observe that the
library is written in an idiomatic functional style, with no special
reasoning principles needed to realize desired functionality.  As long
as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior can become significantly harder to understand.  In
particular, a distributed implementation might wish to
\emph{replicate} the counter state on each node to improve response
time or fault tolerance.  Unfortunately, adding replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose significant performance penalties.  But, without such
heavyweight mechanisms, applying an \C{Add} operation on one node may
not be instantaneously witnessed on another, which may be in the process of
simultaneously attempting to perform its own \C{Add} or \C{Mult}
action.  While synchronizing the activities of all nodes to ensure at
most one such operation is performed at a time is impractical,
designating a single node to hold the counter state eschewing
replication altogether (as in a client-server
configuration~\cite{Armstrong}), is also an undesirable solution,
given the sensitivity of such architectures to network partitions and server failures, and
the negative performance impact it incurs in geo-distributed
environments~\cite{Walter}.  Removing coordination altogether by simply
replicating counter state without having any supporting consistency
protocol is an equally infeasible approach.

@@

nonetheless witness anomalous states that violate the invariant

@@

Framing replication as merging leads to a counter implementation that
bears strong similarity to the original sequential one:
  \begin{ocaml}
    module Replicated_Counter = struct
      include Counter
      let merge lca v1 v2 =
         lca + (v1 - lca) + (v2 - lca)
    end
  \end{ocaml}
The role of \C{lca} (least common ancestor) here captures salient
history - the state resulting from the merge of two versions derived
from the same ancestor state should not unwittingly duplicate the
contributions of the ancestor.  This interpretation of a replicated
datatype is thus given in terms of the evolution of a program state
implicitly associated with the different nodes that comprise a
distributed application with merge operations serving to communicate
and reconcile different local states.

@@

The remainder of the paper is organized as follows.  We present the
\name\ programming model informally in the next section.
Sec.~\ref{sec:mergeable_types} develops a series of examples that
illustrate how merge operations can be defined and used.
Sec.~\ref{sec:formalization} defines an operational semantics, and
formalizes our intuition of a mergeable type in terms of morphisms over
datatype operations.  We also present soundness (consistency) and
progress guarantees enjoyed by the
semantics. Sec.~\ref{sec:system-model} considers an instantiation of
the model suitable for distributed environments with network
partitions and failures.  Implementation details describing how we
incorporate mergeable types into OCaml are presented in
Sec.~\ref{sec:impl}.  Experimental results and benchmarks are given in
Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related}
and~\ref{sec:conclusions} describes related work and conclusions.

### 2017-02-28 paper/meta.tex

\begin{lemma} [\bfseries Non-Monotonicity of Mergeability]
In a legal branching history $H$ produced by the operational
semantics, if two branches, $b_i$ and $b_j$ are not mergeable (as per
Def.~\ref{def:mergeability}), then there exist a series of merges
involving only the mergeable branches that can be performed on $H$ to
yeild a new history $H'$, where $b_i$ and $b_j$ are now mergeable.
\end{lemma}

>>>>

\begin{lemma} [\bfseries Non-Monotonicity of Mergeability]
In a legal branching history $H$ produced by the operational
semantics, if two branches, $b_i$ and $b_j$ are not mergeable (as per
Def.~\ref{def:mergeability}), then there exist a series of merges
involving only the mergeable branches that can be performed on $H$ to
yield a new history $H'$, where $b_i$ and $b_j$ are now mergeable.
\end{lemma}

### 2017-02-28 paper/motivation.tex
edits

In this section, we motivate mergeable data types and \name
programming model via a series of examples.

@@

The counter implementation of Fig.~\ref{fig:counter-adt} serves the
application well as long as it is used in-memory on a single machine.
However, as the application grows in complexity and scales beyond a
single machine, its state may also need to be replicated and
persisted. Unfortunately, replication doesn't come for free, and
existing techniques require significant re-engineering of the
application and its libraries to support replication. A method often
adopted~\cite{crdts, pldi15, gotsman-popl16} is to re-define the operations to
return \emph{effects} instead of values.  An \emph{effect} is a tag that
identifies the operation to be applied uniformly at all
replicas to incorporate the effects of the original operaton. For
instance, \C{Counter.add x v} operation returns \C{Add x} effect,
which, when applied at a replica (see \C{apply}), adds \C{x} to the 
local counter value.  Fig.~\ref{fig:counter-rdt} shows the
\C{Counter} library with operations re-defined to returns effects.
Such re-engineering has to be done carefully, lest an operation may
may end up causing an unintended effect at a remote replica. For
instance, one may incorrectly define \C{mult x v} operation as
generating a \C{Mult x} effect, for it is easy to overlook the fact
that \C{apply (Mult x)} affects different replicas differently
depending on the counter value. Thus, reasoning about the correctness
of effects often involves appeals to system-specific artifacts, such
as replicas, which don't normally manifest in the program.  While
checking the effects (or rather their application via \C{apply}) for
commutativity may help catch some of these oversights, commutativity
is a non-local property, and automatically verifying it is difficult.
Besides introducing the possibility of new bugs, re-engineering
libraries to return effects has obvious software engineering problems
in the sense that it breaks backwards compatibility with the existing
code.

Another downside with effect-generating replicated data type libraries
is that they don't compose. For instance, let us say that an
application wants to use two replicated counters, $c_1$ and $c_2$,
bound by an invariant $c_2 \ge c_1$, which the application duly
enforces when updating $c_2$ and $c_1$. Nonetheless, it may still
witness anamolous states that violate the invariant because updates to
$c_1$ and $c_2$ are applied independently in any order. For example,
if a replica increments both the counters by 1 each, the \C{read} at
another replica, where $(c_1,c_2) = (2,2)$ may witness the increment
to $c_1$, but not $c_2$, thus violating the invariant. The intention
to atomically apply both the effects cannot be captured in the absence
of external mechanisms to compose effects (such as effect-based
transactions~\cite{pldi15}).

Persistence adds another dimension to the problem. Functional data
structures are often large in-memory linked structures that admit
functional updates by creating newer versions that share most of their
structure with previous versions. Sharing is the key to efficiency,
without which every functional update takes time that is at least
linear in the size of the data structure. Unfortunately, the
straightforward way to persist a data structure on disk in a
machine-agnostic format (e.g., JSON) requires serialization, which is
linear in the size of the data structure. An alternative is to
maintain a separate on-disk copy of the data and administer it via a
database system. The downside is that the programmer is now required
to reason in terms of a lower-level abstraction (e.g., a relational
data model) in order to maintain coherence between in-memory and
on-disk representations. Safety guarantees over the in-memory data do
not carry over to the disk, which introduces additional complications
and the possibility of new bugs. Thus, adding persistence to
applications while preserving high-level safety gurantees offered by
the language abstraction is a non-trivial problem.

>>>>

In this section, we motivate mergeable data types and the \name
programming model via a series of examples.

@@

The counter implementation of Fig.~\ref{fig:counter-adt} serves the
application well as long as it is used in-memory on a single machine.
However, as the application grows in complexity and scales beyond a
single machine, its state may also need to be replicated and
persisted. Unfortunately, replication doesn't come for free, and
existing techniques require significant re-engineering of the
application and its libraries to support replication. A method often
adopted~\cite{crdts, pldi15, gotsman-popl16} is to re-define the operations to
return \emph{effects} instead of values.  An \emph{effect} is a tag that
identifies the operation to be applied uniformly at all
replicas to incorporate the effects of the original operation. For
instance, \C{Counter.add x v} operation returns \C{Add x} effect,
which, when applied at a replica (see \C{apply}), adds \C{x} to the 
local counter value.  Fig.~\ref{fig:counter-rdt} shows the
\C{Counter} library with operations re-defined to returns effects.
Such re-engineering has to be done carefully, lest an operation may
may end up causing an unintended effect at a remote replica. For
instance, one may incorrectly define \C{mult x v} operation as
generating a \C{Mult x} effect, for it is easy to overlook the fact
that \C{apply (Mult x)} affects different replicas differently
depending on the counter value. Thus, reasoning about the correctness
of effects often involves appeals to system-specific artifacts, such
as replicas, which don't normally manifest in the program.  While
checking the effects (or rather their application via \C{apply}) for
commutativity may help catch some of these oversights, commutativity
is a non-local property, and automatically verifying it is difficult.
Besides introducing the possibility of new bugs, re-engineering
libraries to return effects has obvious software engineering problems
in the sense that it breaks backwards compatibility with the existing
code.

Another downside with effect-generating replicated data type libraries
is that they don't compose. For instance, let us say that an
application wants to use two replicated counters, $c_1$ and $c_2$,
bound by an invariant $c_2 \ge c_1$, which the application duly
enforces when updating $c_2$ and $c_1$. Nonetheless, it may still
witness anomalous states that violate the invariant because updates to
$c_1$ and $c_2$ are applied independently in any order. For example,
if a replica increments both the counters by 1 each, the \C{read} at
another replica, where $(c_1,c_2) = (2,2)$ may witness the increment
to $c_1$, but not $c_2$, thus violating the invariant. The intention
to atomically apply both the effects cannot be captured in the absence
of external mechanisms to compose effects (such as effect-based
transactions~\cite{pldi15}).

Persistence adds another dimension to the problem. Functional data
structures are often large in-memory linked structures that admit
functional updates by creating newer versions that share most of their
structure with previous versions. Sharing is the key to efficiency,
without which every functional update takes time that is at least
linear in the size of the data structure. Unfortunately, the
straightforward way to persist a data structure on disk in a
machine-agnostic format (e.g., JSON) requires serialization, which is
linear in the size of the data structure. An alternative is to
maintain a separate on-disk copy of the data and administer it via a
database system. The downside is that the programmer is now required
to reason in terms of a lower-level abstraction (e.g., a relational
data model) in order to maintain coherence between in-memory and
on-disk representations. Safety guarantees over the in-memory data do
not carry over to the disk, which introduces additional complications
and the possibility of new bugs. Thus, adding persistence to
applications while preserving high-level safety guarantees offered by
the language abstraction is a non-trivial problem.

### 2017-02-28 paper/semantics.tex

Reduction rules let expression evaluation take a step by rewriting the
expression and suitably updating the branch history ($H$).
\rulelabel{E-App} is the standard beta reduction rule.  The
\rulelabel{E-Run} rule is applicable only when $H$ is empty, i.e.,
when no prior branching structure exists, and commences a distributed
computation with a corresponding version tree. The rule rewrites the
$\C{run}\;v\;s$ expression to $s$, while creating a new branching
structure with two branches: a \emph{top} branch that has just the
initial version (tagged with \C{INIT}), and a branch for the current
thread ($t$) forked-off from the top branch.  The first version on the
current branch ($H(t)$) denotes the same value ($v$) as the initial
version on the top branch, although versions themselves are deemed
distinct. The new version is tagged with a \C{FORK} tag that keeps the
record of its orgin, namely the \C{fork} operation and the branch from
which the current branch is forked. The \rulelabel{E-Fork} rule forks
a new thread with a fresh id ($t'$) and adds it to the thread pool.
The corresponding branch ($H(t')$) is forked from the parent thread's
branch ($H(t)$). The semantics of branch forking is the same as
described above. The \C{fork} expression in the parent thread
evaluates to \C{()}. The \rulelabel{E-Push} rule creates a new version
on the current branch ($H(t)$) using the pushed value ($v$).  Although
our semantics does not directly expose heap-allocated values, the
intention is that $v$ is a replicated object, manipulated on the local
heap that, after push, now becomes subject to merging and coordination
with other replicas.

>>>>

Reduction rules let expression evaluation take a step by rewriting the
expression and suitably updating the branch history ($H$).
\rulelabel{E-App} is the standard beta reduction rule.  The
\rulelabel{E-Run} rule is applicable only when $H$ is empty, i.e.,
when no prior branching structure exists, and commences a distributed
computation with a corresponding version tree. The rule rewrites the
$\C{run}\;v\;s$ expression to $s$, while creating a new branching
structure with two branches: a \emph{top} branch that has just the
initial version (tagged with \C{INIT}), and a branch for the current
thread ($t$) forked-off from the top branch.  The first version on the
current branch ($H(t)$) denotes the same value ($v$) as the initial
version on the top branch, although versions themselves are deemed
distinct. The new version is tagged with a \C{FORK} tag that keeps the
record of its origin, namely the \C{fork} operation and the branch from
which the current branch is forked. The \rulelabel{E-Fork} rule forks
a new thread with a fresh id ($t'$) and adds it to the thread pool.
The corresponding branch ($H(t')$) is forked from the parent thread's
branch ($H(t)$). The semantics of branch forking is the same as
described above. The \C{fork} expression in the parent thread
evaluates to \C{()}. The \rulelabel{E-Push} rule creates a new version
on the current branch ($H(t)$) using the pushed value ($v$).  Although
our semantics does not directly expose heap-allocated values, the
intention is that $v$ is a replicated object, manipulated on the local
heap that, after push, now becomes subject to merging and coordination
with other replicas.
